Parsing Postfix log files

### 2008-01-16 doc/logparser.tex
Added sections on:     Emergent behaviour     Adding new actions     Benefits of storing rules in the database     Ordering of the additional complications Extended and improved the conclusion.

XXX WHAT IS THE CONCLUSION???  FIRST THING IS TO COMPARE TO OTHER PARSERS
REVIEWED

@@

@@

@@

    \item A flow chart showing common paths a mail can take through the
        algorithm.

@@

@@

@@

@@

This section has presented the parsing algorithm, starting with a high
level overview of the basic algorithm, and the first group of difficulties
thwarted.  The revised algorithm was detailed, preceded by a flow chart
depicting the paths a mail may take through the algorithm, and followed by
the actions which are available to rules were described in full.  The
section concludes with the remaining complications discovered and overcome
in the process of refining and completing the algorithm.  Detecting,
diagnosing and defeating complications forms the largest single portion of
this document, mirroring the development of the parser,  The complications
are described in the order they were overcome, with subsequent problems
affecting fewer mails (often by an order or magnitude), though the time
required to solve problems increased with each successive problem.

@@

@@

Enabling users to easily extend the parser to deal with new log lines is a
design imperative, and despite the resulting increase in complexity the job
is quite tractable, though it does raise interesting efficiency and
optimisation questions (answered in section~\refwithpage{rule efficiency}).
For the parser to be useful to the average user, parsing new log lines
produced by Postfix must be as simple a task as configuring Postfix to
produce those log lines was in the first place, otherwise users won't make
to effort to utilise the parser.  Providing a tool to ease the generation
of \regexes{} from unparsed log lines should greatly help users
add rules to parse unparsed lines.

>>>>

XXX USE \textsection{} INSTEAD OF SECTION WHEN REFERRING TO \\OTHER
SECTIONS\@?

@@

An important but easily overlooked benefit of storing the rules in the
database is the connection between rules and results: if more information
is required when examining a result, the rules which produced the database
entries are available for inspection (and each result references the rule
which created it).  There is no ambiguity about which rule resulted in a
particular result, eliminating one potential source of confusion.

@@

\label{creating new rules}

@@

    \item A flow chart showing common paths a mail can take through the
        algorithm, and discussing the emergent behaviour observed in the
        parser.

@@

The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
which is far more complicated than the behaviour described by the
algorithm.  The top-level parser could be written in pseudo-code like so
(with indentation level indicating flow control):

\begin{verbatim}
for each line in the input files:
    for each rule defined by the user:
        if this rule matches the input line:
            perform the action specified by the rule
            skip the remaining rules
            process the next input line
    warn the user that the input line was not parsed
\end{verbatim}

Yet despite the brevity of the algorithm above the (simplified) flow chart
below contains twenty one states, three entry points, five exit points,
five loops, four explicit branches, and eight implicit branches.  Several
more states and corresponding branches are omitted for the sake of clarity
(these extra states are caused by solving the difficulties described in
section~\refwithpage{additional complications}).

The solution to the second last complication encountered
(section~\refwithpage{out of order log lines}) requires that a list of the
acceptable combinations of Postfix programs a mail can pass through;
however this does not correspond to the \textit{actions} a mail must pass
through.  Before that complication was overcome, i.e.\ for the majority of
the development of this parser, there was nothing in the parser explicitly
encoding the states or paths a mail could take; the vast majority of the
parser has no need for this flow of data to be specified.  This eases the
process of adding new actions (described in section~\refwithpage{adding new
actions}), as there is no requirement to explicitly insert the action into
an algorithmic version of the flow chart below.

@@

\subsection{Adding new actions}

\label{adding new actions}

Adding new actions is almost as simple as adding new rules, though actions
are not stored in the database.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action by calling the parser method add\_actions() with the subroutine
name as a parameter.  No other work is required on the implementor's part;
all of their attention and effort can be focused on the correctness of
their action.  The only negative aspect is that the process involves
editing the parser source code, which makes upgrading to a later version of
the parser more difficult, though by no means impossible.  If the author of
the new action wishes they can take advantage of the parser's object
oriented implementation~\cite{Wikipedia-object-orientation} by subclassing
it and implementing their changes in the derived class, allowing future
upgrading of the parser with greatly reduced chance of conflicts.  The
action may need to extend the list of valid combinations described in
section~\refwithpage{out of order log lines} if the addition creates a
different set of acceptable programs, but this is unlikely to occur, as it
would require parsing log lines from Postfix components the parser
currently ignores.\footnote{The mail server used for development does not
utilise the \daemon{lmtp} or \daemon{virtual} delivery agents, so this
parser does not deal with those components.  Adding new rules to parse
their log lines is a simple process, though if their behaviour differs
significantly from the \daemon{smtp} or \daemon{local} delivery agents new
actions may be required.}

@@

The complications described in this section are listed in the order in
which they were encountered during development of the parser.  Each of
these complications caused the parser to operate incorrectly, generating
either warning messages or leaving mails in the state table.  The frequency
of occurrence is much higher at the start of the list, with the first
complication occurring several orders of magnitude more frequently than the
last.  When deciding which problem to address next, the most common was
always chosen, as resolving the most common problem would yield the biggest
improvement in the parser, and also prune the greatest number of mails from
the state tables and error messages.

@@

This section has presented the parsing algorithm, starting with a high
level overview of the basic algorithm, and the first group of difficulties
thwarted.  The revised algorithm was detailed, preceded by a flow chart
depicting the paths a mail may take through the algorithm, and followed by
the actions which are available to rules were described in full.  The
parser exhibits emergent behaviour far more complicated than the behaviour
encoded in the algorithm, and this behaviour is covered in the flow chart
subsection.  The section concludes with the remaining complications
discovered and overcome in the process of refining and completing the
algorithm.  Detecting, diagnosing and defeating complications forms the
largest single portion of this document, mirroring the development of the
parser,  The complications are described in the order they were overcome,
with subsequent problems affecting fewer mails (often by an order or
magnitude), though the time required to solve problems increased with each
successive problem.

@@

XXX PICK A RANDOM SAMPLE OF LOG LINES AND ENSURE THEY ARE PARSED
PROPERLY\@.

@@

Given the diverse nature of the log lines and the ease with which
administrators can cause new lines to be logged
(section~\refwithpage{postfix background}), enabling users to easily extend
the parser to deal with new log lines is a design imperative
(section~\refwithpage{parser design}).  Despite the resulting initial
increase in complexity the job is quite tractable, though it does raise
interesting efficiency and optimisation questions (answered in
section~\refwithpage{rule efficiency}).  Providing a tool
(section~\refwithpage{creating new rules}) to ease the generation of
\regexes{} from unparsed log lines should greatly help users add rules to
parse formerly unparsed lines.

The implementation not only substantially eases the parsing of new log
lines, it makes adding new actions
(section~\refwithpage{parsing-algorithm}) an almost trivial task.  The
simplicity of adding a new action (section~\refwithpage{adding new
actions}) frees the implementor from worrying about how their action might
disrupt parsing of other lines or the behaviour of other actions, allowing
their concentration to be focused on correctly implementing their action.

The division of the parser into rules and actions is unusual because it
divides the two so completely; although parsers are often divided (lex and
yacc~\cite{lex-and-yacc} based parsers being an obvious example), the parts
are usually quite internally interdependent, and will be combined into a
complete parser by the compilation process; in contrast \parsername{} keeps
the rules and actions separate until the parser runs.  XXX EXTEND THIS\@;
SOMETHING ABOUT THE CONTROL FLOW BEING DIFFERENT, I.E. RULE DETERMINES
ACTION, RATHER THAN ACTION EXPECTING INPUT\@.

The emergent behaviour~\cite{Wikipedia-Emergence} exhibited by the rules
and actions is also interesting, and is discussed beside the flow chart
depicting said emergent behaviour (section~\refwithpage{flow-chart}).  This
emergent behaviour greatly eases
